Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-16 submission/manuscript.Rmd
begin discussion edits

Although many germline genes have shared functions in the male and female germline, some have unique or sex-biased expression. Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females would primarily express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H). We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This criteria yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes. We found regardless of sex, egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J). Furthermore, germline genes dysregulated exclusively in either male or female mutants were also not biased towards their corresponding germ cell sex (Figure 4I). Altogether, these results demonstrate sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation via ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled derepressive mechansisms governing distinct classes of germline gene in somatic lineages. Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. KDM5C's Y-chromsome homolog, KDM5D, exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007] and it is currently unknown to regulate germline gene expression. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.

We found KDM5C is largely dispensible for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed at the 2-cell stage and in na&#239;ve ESCs for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. Although expressed in na&#239;ve ESCs, "self-renewal" germline genes like _Dazl_ are silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs also express _Dazl_, they did not express 2-cell-specific genes. These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs), rather than de-differentiating into 2-cell-like cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated. Thus, ectopic germline gene expression occurs along with proper somatic development in _Kdm5c_-KO animals.

Our work provides novel insight into the cross-talk between H3K4me and CpGme. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_ and _Naa11_, strongly impaired CGI methylation, and resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 primarily do not colocalize with CpGme[@weberDistributionSilencingPotential2007] and can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Together, this suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007]. Furthermore, these results indicate the requirement for catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes.

By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct derepressive mechanisms governing early versus late-stage germline developmental programs. Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation and acts as break against runaway activation of germline-specific programs.

The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle, liver, and even ovary-biased transcripts within the male _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_. Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023]. Intriguingly, mouse and human models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary. Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

>>>>

Although many germline genes have shared functions in the male and female germline, e.g. PGC formation,  meiosis, and genome defense, some have unique or sex-biased expression. Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females would primarily express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H). We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This criteria yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes. We found regardless of sex, egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J). Furthermore, germline genes dysregulated exclusively in either male or female mutants were also not biased towards their corresponding germ cell sex (Figure 4I). Altogether, these results demonstrate sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.

@@

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We first characterized tissue-enriched genes expressed within the mouse _Kdm5c_-KO brain and identified substantial derepression of testis, liver, muscle, and ovary-enriched genes. Testis genes significantly enriched within the _Kdm5c_-KO amygdala and hippocampus are specific to the germline and not expressed within somatic cells. _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly express key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the adult brain primarily expresses genes important for late spermatogenesis. We demonstrated that although _Kdm5c_ mutant sex did not influence whether sperm or egg-specific genes were misexpressed, female EpiLCs are more sensitive to germline gene de-repression. Germline genes can become aberrantly expressed in _Kdm5c_-KO cells via an indirect mechanism, such as activation through ectopic RFX transcription factors. Finally, we found KDM5C is dynamically regulated during ESC to EpiLC differentiation to promote long-term germline gene silencing through DNA methylation at CpG islands. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression, we unveiled unique mechansisms governing the misexpression distinct germline gene classes within somatic lineages. Furthermore, these data provide molecular footholds which can be exploited to test the overarching contribution of ectopic germline gene expression upon neurodevelopment.

By comparing _Kdm5c_ mutant males and females, we revealed germline gene supression is sexually dimorphic. Although sex did not impact whether egg or sperm-specific genes were dysregulated, organismal sex did greatly influence the degree of germline gene dysregulation. Female EpiLCs are more severely impacted by loss of KDM5C-mediated germline gene suppression, yet this difference is not due to the increased number of germline genes on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. Increased female sensitivity to germline gene de-repression may be related to females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Males could partially compensate for KDM5C's loss via the Y-chromsome homolog, KDM5D, which exhibits weaker demethylase activity than KDM5C[@iwaseXLinkedMentalRetardation2007]. However, KDM5D has not been reported to regulate germline gene expression. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to elucidate the biological ramifications and underlying mechanisms.

We found KDM5C is largely dispensible for promoting normal gene expression during development, yet is critical for suppressing ectopic developmental programs. It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes like _Dazl_ are also expressed at the 2-cell stage and in na&#239;ve ESCs/inner cell mass for their role in pluripotency and self-renewal[@suzukiLossMAXResults2016; @torres-padillaTransposonsTotipotency2020;@yangChemicalinducedChromatinRemodeling2022; @guptaGenomewideScreenReveals2023]. These "self-renewal" germline genes are then silenced during ESC differentiation into epiblast stem cells/EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. We found that while _Kdm5c_-KO EpiLCs express _Dazl_, they did not express 2-cell-specific genes like _Zscan4c_. These data suggest the 2-cell-like state reported in _Kdm5c_-KO ESCs[@guptaGenomewideScreenReveals2023] likely reflects KDM5C's primary role in germline gene repression. Germline gene misexpression in _Kdm5c_-KO EpiLCs may indicate they are differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011; @magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. Yet, _Kdm5c_-KO EpiLCs had normal cellular morphology and properly expressed markers for primed pluripotency, including _Otx2_ which blocks EpiLC differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. In addition to unimpaired EpiLC differentiation, _Kdm5c_-KO gross brain morphology is overall normal[@iwaseMouseModelXlinked2016] and hardly any brain-specific genes were significantly dysregulated. Thus, ectopic germline gene expression occurs along with overall proper somatic differentiation in _Kdm5c_-KO animals.

Our work provides novel insight into the cross-talk between H3K4me and CpGme, which are often mutually exclusive[@weberDistributionSilencingPotential2007]. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_ and _Naa11_, strongly impaired CGI methylation, and resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. This suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme[@longProtectionCpGIslands2016; @weberDistributionSilencingPotential2007]. Furthermore, these results indicate the requirement for catalytic activity can change depending upon the locus and developmental stage, even for the same class of genes. Further experiments are required to determine if catalytically inactive KDM5C can suppress germline genes at later developmental stages.

By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct derepressive mechanisms governing early versus late-stage germline developmental programs. Previous work on germline gene silencing has focused on genes with promoter CGIs[@borgelTargetsDynamicsPromoter2010; @weberDistributionSilencingPotential2007], and indeed the majority of KDM5C targets in EpiLCs were germ cell identity genes harboring CGIs. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_, _Stra8_, and _Dazl_ turn on downstream germline programs, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation to act as a break against runaway activation of germline-specific programs.

The above work provides the mechanistic foundation for KDM5C-mediated repression of tissue and germline-specific genes. However, the contribution of these ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. In addition to germline genes, we also identified significant enrichment of muscle and liver-biased transcripts within the _Kdm5c_-KO brain. Intriguingly, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022]. _APOC1_ dysregulation is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010] and overexpression of _Apoc1_ in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008]. KDM5C may therefore be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes like _Apoc1_.

Given germline genes have no known functions within the brain, their impact upon neruodevelopment is currently unknown. Ectopic testicular germline transcripts have been observed in a variety of cancers[@nielsenEctopicExpressionTestis2016; @adebayobabatundeCancerTestisGenes2017], including brain turmors in _Drosophila_ and mammals and shown to promote cancer progression[@janicEctopicExpressionGermline2010; @ghafouri-fardExpressionCancerTestis2012; @ninBiologyCancerTestisAntigens2023]. Intriguingly, mouse and human models for other chromatin-linked neurodevelopmental disorders also display impaired soma-germline demarcation[@bonefasSomatogermlineTransformationChromatinlinked2021; @velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014; @schaeferControlCognitionAdaptive2009; @samacoCrhOprm1Mediate2012], such as DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2)[@bonefasSomatogermlineTransformationChromatinlinked2021]. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders with similar erosion of the germline versus soma boundary. Further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.
